Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2023-09-24 manuscript/manuscript.Rmd
results update.

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c). Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation. A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness. Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.

>>>>

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c). Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation. A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness. Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
```

##  Classification Validation

We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs. two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables. All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p< 0.001) which correctly classified 93.8% of artifacts. We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses. There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746). Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes. Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core. This piece, typologically classified as a “knife”, is the smallest (93mm) retouched flake in the Acheulean sample. When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts. Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs. 186g, Cohen’s d=1.137, p<0.001) and less heavily reduced (mean cSDI 0.057 vs. 0.103, Cohen’s d = -0.884, p < 0.001) despite having similar FAI (mean 0.52 vs. 0.46, Cohen’s d=0.271, p=0.266).

Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e. excluding unmodified large flakes, n = 115) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables. Both shape PCs and FAI were retained, while cSDI was not entered. This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p<0.001) which correctly classified 71.3% of artifacts. Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2). A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2). As an intermediate type, “handaxes” were correctly classified only 43.9% of the time. In agreement with others (Duke, et al., 2021, Kuhn, 2021, Presnyakova, et al., 2018) we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.

##  Effects of Reduction on Shape
To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2). In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p <0.001) and FAI (r=-0.294, p<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p<0.001). However, it is clear that these overall effects conflate different trends in Mode 1 vs. Mode 2 cores, as well as in cores executed on flake vs. pebble/cobble bases. Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.

In Mode 1 cores, a weak negative effect of FAI on PC2(pointedness) (n = 76, $\beta$ =-0.006, r=0.240, p= 0.037) suggests that increasing extent of modification tends to slightly increase pointedness (Figure 6a). No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance. FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).

In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form. In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d). Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406). Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p < 0.001). Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e. both upper and lower limits of cSDI increase with FAI).
Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness). In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs. 0.774 and 0.693 vs. 0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).

A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined. These heavily worked “indeterminate” cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x). They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).
